---
title: "BUDA 525 Final— Group Assignment"
author: "Brady Beckner, Ian Bush, Dylan Cunningham"
date: "9/28/2026"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Directions

This is the final group project for BUDA 525.  Submit both the .Rmd and word document to the appropriate link on ecampus by 11:59 on Thursday October 8, 2025.  There are 4 questions on this assignment.  The first two questions are discussion type questions which require no data analysis but do require substantial insights into the material that has been covered.  The final two questions are data analysis type questions which require you to analyze data and communicate relevant findings.  You should only show results in this document that relevant to your discussions. Remember you can use ECHO=FALSE, and other R markdown arguments to present your code.  Since I have the .Rmd I can see all code that was not presented in the word document. 


## Problem 1 (20 points)

Compare and contrast the bootstrap and randomization test procedures we have discussed.  In this discussion you should describe the purpose of each method (why would you use them), the procedure they follow, and then how they differ. Finally give a few sentences on each that would explain to a novice what these methods do and why we should trust their results.

#Bootstrap and randomization testing both use the data that we have available to get a deeper understanding of a statistical problem. Bootstrap is typically used to see how much a statistic such as the mean or median changes from sample to sample. Bootstrap does this by taking samples from the original dataset with replacement and calculating the statistic each time. A randomization test is used when testing whether there is a difference or relationship between groups. Randomization tests change the group labels and calculate the statistic multiple times to create a distribution based on the assumption of the null hypothesis being true. You can then compare the actual result to the distribution and calculate the p-value.

	#The main difference between bootstrap and randomization tests is that bootstrap looks at how much the statistic could vary, and randomization tests look at whether the result could have happened by chance if the null hypothesis were true. For a novice, you can think of a bootstrap as repeatedly creating new versions of your sample to see how much your answer changes. A randomization test is mixing up the groups to see how often we would get a result similar to the one we actually found. You can have confidence in these methods because they repeat the process many times, which gives a better idea of the differences in results we could expect instead of relying on just one sample or calculation.


## Problem 2 (20 Points)

Compare and contrast the model selection procedures we've discussed in this course.  Specifically focus your discussion on R-squared, Anova, AIC, BIC, K-Fold Cross Validation, and Random Splitting.  You should further focus on when are these methods able to be used and when are they not.  Finally discuss your preference for  a technique and give reasons why.


#The model selection procedures we have used in this course help us decide which model is better, but they work in different ways.  R-squared shows us how much of the variation in the response is explained by the model. A higher R-squared typically means the model fits the data better, but it can also go up when more variables are added, so it does not immediately mean the model is better. ANOVA can be used to compare nested models and see if adding variables makes any real difference. AIC and BIC look at how well the model fits and also consider how many variables are in the model. For both AIC and BIC, a lower value is typically better. These methods work best when comparing models using the same data and tend to focus more on model fit than on data prediction.

	#K-Fold Cross Validation and Random Splitting focus on how well a model will predict new data. K-Fold Cross Validation splits the data into multiple groups, and the model is trained on some of the groups and tested on the groups that are not included. This repeats until each group is used for testing. Random Splitting is similar, but the training and testing groups are randomly selected, and the process allows for it to be repeated. These are useful when the goal is prediction, but they may not work as well with data that has a certain order. We prefer K-Fold Cross Validation because it gives a good understanding of how the model will perform on new data and uses the observation data from both the training and the testing.


## Problem 3 (30 Points)

In the `Credit` data in the `ISLR` package it contains 400 customers and information on their credit history.  For full information of the data look at the `help` file.  A company has approached us to better understand factors that influence the `Balance` variable which is average credit card balance in USD.  Using the information in the model discuss the influential factors, and discuss the factors you choose to put in the model.  Do you have any concerns about the use of certain variables in the model? Discuss how your model was created and any insights you can provide based on the results.  HINT:  Adding Gender and/or Ethnicity could be controversial or illegal in some uses of this this model you should discuss your decision on these variables and how it effects the organizations ability to use your model for prediction or inference.  


```{r}
library(ISLR)
head(Credit)
```


## Problem 4 (30 Points)

The `Salaries` data in the `carData` package contains information on academic salaries in 2008 and 2009 in a college in the US. A data dictionary can be found in the `help` file for the data. This data was collected as part of an on-going effort of the college to monitor salary differences between male and female faculty members.  We have been asked to investigate the gender gap in the data, but also what other information that may be relevant to administrators (i.e. salary growth for years of service, discipline based growth, etc).  Investigate if there is a gender gap, but also provide insights on other drivers that you may see of salary in the data.  Is your model suitable to make offers based on the information provided?  Explain your reasoning.  Provide insights into any other information you find of interest.  
```{r}
library(carData)
head(Salaries)
```
